One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-06 paper/appendix.tex
update

\section{The background torque density in a three-dimensional disc with
  a fixed temperature profile}\label{tbg_deriv}
We give a brief derivation of the angular momentum exchange between
linear perturbations and the background disc. We consider a
three-dimensional disc in which the equilibrium pressure and density
are related by 
\begin{align}\label{iso_cond}
  p = c_s^2(R,z)f(\rho),
\end{align} 
where $f(\rho)$ is an arbitrary function of $\rho$ with dimensions of
mass per unit volume, and $c_s$ is a prescribed function of
position with dimensions of velocity squared. The equilibrium disc
satisfies 
\begin{align}
  R\Omega^2(R,z) &= \frac{1}{\rho}\frac{\p p}{d R} +
  \frac{\p\Phi_\mathrm{tot}}{\p R}, \\
  0 &= \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_\mathrm{tot}}{\p
    z}. 
\end{align}
Note that, in general, the equilibrium rotation $\Omega$ depends on
$R$ and $z$.

@@

So far we have not invoked an energy equation. For 
adiabatic perturbations $T_\mathrm{BG}$ is zero
\citep{lin93b}. However, if we impose the equilibrium relation
Eq. \ref{iso_cond} to hold in the perturbed state, then
\begin{align}
  \delta p = c_s^2(R,z) f^\prime \delta\rho,
\end{align}
where $f^\prime = df/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
obtain
\begin{align}\label{tbg_3d_2}
  T_\mathrm{BG} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\delta\rho\bm{\xi}^*\cdot\nabla c_s^2 +
    \xi_R^*\xi_z \left(\frac{\p\rho}{\p z}\frac{\p c_s^2}{\p R} -
      \frac{\p\rho}{\p R}\frac{\p c_s^2}{\p z}\right)\right],
\end{align}
where the equilibrium equations were used. 
At this point setting $\xi_z=0$ gives $T_\mathrm{BG}$ for
perturbations with no vertical motion, 
\begin{align}
  T_\mathrm{BG,2D} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left(\delta\rho \xi_R^*  \p_R c_s^2 \right),
\end{align}
and is equivalent to the 2D
expression, Eq. \ref{baroclinic_torque}, with $\delta\rho $ replaced
by $\delta \Sigma$ and $p=c_s^2\rho$. % The 2D expression is convenient
% for razor-thin discs or perturbations without vertical motion because
% one can measure $\delta\rho$ (or $\delta \Sigma)$ directly.

>>>>

\section{The background torque density in a three-dimensional disc with
  a fixed temperature profile}\label{tbg_deriv}
We give a brief derivation of the angular momentum exchange between
linear perturbations and the background disc. We consider a
three-dimensional disc in which the equilibrium pressure and density
are related by 
\begin{align}\label{iso_cond}
  p = c_s^2(R,z)F(\rho),
\end{align} 
where $F(\rho)$ is an arbitrary function of $\rho$ with dimensions of
mass per unit volume, and $c_s$ is a prescribed function of
position with dimensions of velocity squared. The equilibrium disc
satisfies 
\begin{align}
  R\Omega^2(R,z) &= \frac{1}{\rho}\frac{\p p}{d R} +
  \frac{\p\Phi_\mathrm{tot}}{\p R}, \\
  0 &= \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_\mathrm{tot}}{\p
    z}. 
\end{align}
Note that, in general, the equilibrium rotation $\Omega$ depends on
$R$ and $z$.

@@

So far we have not invoked an energy equation. For 
adiabatic perturbations $T_\mathrm{BG}$ is zero
\citep{lin93b}. However, if we impose the equilibrium relation
Eq. \ref{iso_cond} to hold in the perturbed state, then
\begin{align}
  \delta p = c_s^2(R,z) F^\prime \delta\rho,
\end{align}
where $F^\prime = dF/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
obtain
\begin{align}\label{tbg_3d_2}
  T_\mathrm{BG} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left[\delta\rho\bm{\xi}^*\cdot\nabla c_s^2 +
    \xi_R^*\xi_z \left(\frac{\p\rho}{\p z}\frac{\p c_s^2}{\p R} -
      \frac{\p\rho}{\p R}\frac{\p c_s^2}{\p z}\right)\right],
\end{align}
where the equilibrium equations were used. 
At this point setting $\xi_z=0$ gives $T_\mathrm{BG}$ for
perturbations with no vertical motion, 
\begin{align}
  T_\mathrm{BG,2D} = -\frac{m}{2}\frac{p}{\rho
    c_s^2}\imag\left(\delta\rho \xi_R^*  \p_R c_s^2 \right),
\end{align}
and is equivalent to the 2D
expression, Eq. \ref{baroclinic_torque}, with $\delta\rho $ replaced
by $\delta \Sigma$ and $p=c_s^2\rho$. % The 2D expression is convenient
% for razor-thin discs or perturbations without vertical motion because
% one can measure $\delta\rho$ (or $\delta \Sigma)$ directly.

### 2014-11-06 paper/discussion.tex
backup
update

Of course, we have been assuming a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so the above can only be
considered as an order-of-magnitude estimate. Nevertheless, the
expected growth rate is in reasonable agreement with numerical
results.

@@

\subsubsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11a}.

@@

%speculations - long term balance. forced temp gradient drives
%instability (disc gains ang mom, spiral loses ang mom). but if spiral
%shocks and dissiple, it dumps negative angular momentum (disc gains
%negative ang mom). balance between gain from forced temp gradient and
%loss due to dissiplatoin

>>>>

Of course, we have been assuming a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
Eq. \ref{theoretical_rate} is only an order-of-magnitude
constraint. Nevertheless, the xpected growth rate is in reasonable
agreement with numerical results.

%also total torque is integral -> may get cancelations

@@

\subsubsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}.

@@

%required disc structure (for confined SG spiral)

\subsection{Speculations}
%speculations - long term balance. forced temp gradient drives
%instability (disc gains ang mom, spiral loses ang mom). but if spiral
%shocks and dissiple, it dumps negative angular momentum (disc gains
%negative ang mom). balance between gain from forced temp gradient and
%loss due to dissiplatoin

%other source of m=1 disturbance -> eccentric modes (adams, PJ06,
%hopkins, tremaine)
